### 2021-11-15 documents/sections/extensions.tex
tiriyoan *t- review

\subsection{\PTir \rc{t-}}
\label{sec:taranoan}
The Taranoan languages were already grouped together by \textcite{girard1971proto}, consisting of closely related \trio and \akuriyo (subsumed here under the moniker Tiriyoan), and the more distant member \carijo.
\textcite{meira1998proto} provides an extensive phonological, morphological, and lexical reconstruction of \PTar.
The \setone paradigms of \trio and \akuriyo contain an interesting puzzle: \PC \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} \rc{w-} seem to have switched places.
This resulted in a regular \gl{1}\gl{s_a_} marker of the form \envr{\rc{t͡ʃ-}}{\obj{e}}, \envr{\rc{t-}}{\obj{ə}} \pcref{tab:ptirreg}; the latter allomorph was subsequently replaced by \obj{k-} in \akuriyo \pcref{sec:akuriyo}.

\input{floats/ptirreg}
%This is illustrated by comparing reconstructible \PTir forms with more conservative \wayana \exref{sat} and \kalina \exref{13w}, which can be taken as representative for \PC.
%
%\ex<sat>
%\begin{tabular}[t]{@{}llllll@{}}
%\gl{1}\gl{s_a_} & \PC & \wayana & \PTir & \trio & \akuriyo\\
%\qu{yawn} & \rc{w-e-mɨta-po-} & \obj{w-emtapɨ-} & \rc{t͡ʃ-entapo-} & \obj{s-entapo-} & \obj{t͡ʃ-entapo-}\\
%\qu{bathe} & \rc{w-e-pɨ-} & \obj{w-epɨ-} & \rc{t͡ʃ-epɨ-} & \obj{s-epɨ-} & \obj{t͡ʃ-epɨ-}\\
%\qu{talk} & \rc{w-ət-uru} & \obj{w-əturu-} & \rc{t-əturu-} & \obj{t-əturu-} & \obj{k-əturu-}\footnotemark\\
%\qu{learn} & \rc{w-ət-eme-pa} & \obj{w-əhepa-} & \rc{t-əempa-}\footnotemark & \obj{t-əenpa-} & \obj{k-əempa-}\\
%\end{tabular}\\
%\parencites[28]{camargo2010wayana}[292, 664]{triomeira1999}[77, 86, 82]{gildea1994akuriyo}[50, 208, 213]{wayanatavares2005}
%\xe
%\addtocounter{footnote}{-1}
%\footnotetext{\PTir \rc{t-} was subsequently replaced by \obj{k-} in \akuriyo \pcref{sec:akuriyo}.}
%\addtocounter{footnote}{1}
%\footnotetext{
%The \trio form is not due to dissimilation from \rc{mp} to \obj{np}, but rather a phonemic transcription rendering the syllable-final archiphoneme \pnm{N} as \obj{n} \parencite[31]{triomeira1999}.
%Further, \PTar intervocalic \rc{t͡ʃ} was lost entirely in \trio (except after \rc{i}) and is reflected as \obj{ʔ} in \akuriyo \parencite[31--32]{meira1998proto}.
%Given the original shape of the detransitivizers \rc{ət(e)-} and \rc{e-}, \trio \obj{ə-enpa} \qu{to learn} contains a reflex of the former with loss of \rc{t}, so \akuriyo \obj{əempa} would be expected to have a glottal stop: ?\obj{əʔempa}.
%Its absence could either be due to \trio influence or not being transcribed by \textcite{gildea1994akuriyo}.
%Since there is no evidence for a glottal stop in \PTir (\rc{əʔempa}), I do not reconstruct it.}
%
%\ex<13w> 
%\begin{tabular}[t]{@{}llllll@{}}
%\gl{1}>\gl{3}& \PC & \kalina & \PTir & \trio & \akuriyo \\
%\qu{see} & \rc{t-ene-} & \obj{s-ene-} & \rc{w-ene-} & \obj{w-ene-} & \obj{(w-)ene-} \\
%\qu{bring} & \rc{t-enepɨ-} & \obj{s-enepɨ-} & \rc{w-enepɨ-} & \obj{w-enepɨ-} & Ø\obj{-enepɨ-} \\
%\end{tabular}\\
%\parencites[425--426]{hoff1968carib}[28, 304]{triomeira1999}[86--87]{gildea1994akuriyo}
%\xe
%%
%Thus, \wayana preserves the \PC \gl{1}\gl{s_a_} marker \rc{w-}, while it was replaced by \rc{t-} (\phonr{}{\rc{t͡ʃ-}}{\rc{e}}) in \PTir \exref{sat}.
%%Both \qu{to yawn} and \qu{to bathe} are transparently derived \gl{s_a_} verbs, from \rc{mɨta-po} \qu{to open the mouth of} and \rc{pɨ} \qu{to wash}, respectively.
%On the other hand, \gl{1}>\gl{3} scenarios show a reflex of \rc{t-} in \kalina, while they took on \rc{w-} in \PTir \exref{13w}.
%
The question of how this switch happened in detail \parencite[107--112]{meira1998proto} still has no answer, although it seems necessary to assume a scenario whereby both \rc{t-} and \rc{w-} for a time occurred on both transitive and intransitive verbs \parencite[112]{meira1998proto}.\footnote{
In fact, even the issue of \emph{when} this happened is open.
It could have happened at the \PTar stage, but the subsequent introduction of \obj{j-} in \carijo \pcref{sec:carijo} would have erased any traces of such an innovation.
Accordingly, \textcite{meira1998proto} hesitates to assign this extension to a specific proto-language.
We take a conservative stance and reconstruct it to \PTir only, but acknowledge the possibility of it happening already in \PTar.}
%
As for intransitive verbs unaffected by the spread of \rc{t-}, \textcite{meira1998proto} reconstructs the first four items in \cref{tab:ptirverbs} as taking \rc{w-} in \PTar; we offer our reconstructed \PTir forms.
To this list, we add the other copular root \rc{eʔi} \parentext{\PTar \rc{et͡ʃi} \parencite[165]{meira1998proto}}, which has first person \obj{w-}, at least in \trio.

@@

\pex<exp>
\a<aku-164> \akuriyo \parencite[][86]{gildea1994akuriyo}\\
\begingl
\gla (w)i-toka//
\glb \gl{1}>\gl{3}-hit//
\glft \qu{I hit him/her.}//
\endgl
\a<aku-159> \akuriyo \parencite[][114]{meira1998proto}\\
\begingl
\gla kəʔ-eepɨ//
\glb \gl{1+2}-come//
\glft \qu{We came.}//
\endgl
\xe
%
%For \akuriyo \qu{to go}, \textcite{gildea1994akuriyo} registered a different prefix \obj{wɨ-}, rather than \posscite{meira1998proto} \obj{ə-}.\footnote{The \akuriyo recorded by Gildea potentially has strong \trio and/or \wayana influence \parencite[253]{gildea1998}.}
%The two forms can be reconciled by a specific idiosyncratic combination of sound changes:
%We suggest that \obj{ə} is the outcome of the lowering of the \rc{ɨ} in the prefix \rc{wɨ-}; the same has happened to the vowel in the \gl{1+2} prefix \rc{kɨt-} \exref{exp.aku-159}.
%Also, \obj{w-} appears to have been subject to ongoing erosion in \akuriyo, also evidenced by its absence in \qu{to come} and \qu{to be}, but its presence in \qu{to say} \exref{ptirw}.
%This erosion is also found in the etymologically related \gl{1}>\gl{3} prefix \exref{exp.aku-164}, as well as in other Cariban languages, like \ikpeng \pcref{sec:pekodian} or \hixka \pcref{sec:waiwaian}.
%
%
%
For \qu{to come}, \textcite[114--115]{meira1998proto} reconstructs \PTar \rc{əepɨ} for first person, and \rc{eepɨ} for the other person values, based on an idiosyncratic paradigmatic pattern in \trio and the vowel length in \akuriyo.
\akuriyo (and \carijo) then levelled this original distribution, similar to what we have suggested for Pekodian \pcref{sec:pekodian}.
We agree with this scenario, with the exception that \trio \obj{əepɨ} is clearly a reflex of \rc{ət-epɨ} \pcref{sec:come}, meaning that the \PTir form would have been \rc{əʔepɨ}.

@@

In addition, \textcite{gildea1994akuriyo} recorded four more \akuriyo verbs seemingly not affected by innovative \rc{t-} \exref{akumov.aku}, all \obj{e}-initial movement verbs.
We have only found a \trio cognate for \obj{erama} \qu{to return}, which behaves like a regular \gl{s_a_} verb in taking \obj{s-} \exref{akumov.tri}.
Further, these verbs are not mentioned by \textcite{meira1998proto}, who was also working with \posscite{gildea1994akuriyo} data.
Given that this data potentially has strong \trio and/or \wayana influence \parencite[253]{gildea1998} and the lack of support by Meira's data, we cannot reconstruct these verbs as not being affected by the extension of \PTir \rc{t-}.

>>>>

\subsection{\PTir \rc{t-}}
\label{sec:taranoan}
The Taranoan languages were already grouped together by \textcite{girard1971proto}, consisting of closely related \trio and \akuriyo (subsumed here under the moniker Tiriyoan), and the more distant member \carijo.
\textcite{meira1998proto} provides an extensive phonological, morphological, and lexical reconstruction of \PTar.
He faced an interesting puzzle in the \setone paradigms of \trio and \akuriyo: \PC \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} \rc{w-} seem to have switched places.
This resulted in a regular \gl{1}\gl{s_a_} marker of the form \envr{\rc{t͡ʃ-}}{\obj{e}}, \envr{\rc{t-}}{\obj{ə}} \pcref{tab:ptirreg}.%
\footnote{The latter allomorph was subsequently replaced by \obj{k-} in \akuriyo \pcref{sec:akuriyo}.}
\input{floats/ptirreg}
The question of how this switch happened in detail \parencite[107--112]{meira1998proto} still has no answer, although it seems necessary to assume a scenario whereby both \rc{t-} and \rc{w-} for a time occurred on both transitive and intransitive verbs \parencite[112]{meira1998proto}.\footnote{
In fact, even the issue of \emph{when} this happened is open.
It could have happened at the \PTar stage, but the subsequent introduction of \obj{j-} in \carijo \pcref{sec:carijo} would have erased any traces of such an innovation.
Accordingly, \textcite{meira1998proto} hesitates to assign this extension to a specific proto-language.
We take a conservative stance and reconstruct it to \PTir only, but acknowledge the possibility of it taking place already in \PTar.}
%
Regarding \gl{s_a_} verbs unaffected by the spread of \rc{t-}, \textcite{meira1998proto} reconstructs the first four items in \cref{tab:ptirverbs} as taking \rc{w-} in \PTar; we offer our reconstructed \PTir forms.
To this list, the other copular root \rc{eʔi} \parentext{\PTar \rc{et͡ʃi} \parencite[165]{meira1998proto}} can be added, which has first person \obj{w-}, at least in \trio.

@@

\pex<exp>
\a<aku-164> \akuriyo \parencite[][86]{gildea1994akuriyo}\\
\begingl
\gla (w)i-toka//
\glb \gl{1}>\gl{3}-hit//
\glft \qu{I hit him/her.}//
\endgl
\a<aku-159> \akuriyo \parencite[][114]{meira1998proto}\\
\begingl
\gla kəʔ-eepɨ//
\glb \gl{1+2}-come//
\glft \qu{We came.}//
\endgl
\xe
%
%For \akuriyo \qu{to go}, \textcite{gildea1994akuriyo} registered a different prefix \obj{wɨ-}, rather than \posscite{meira1998proto} \obj{ə-}.\footnote{The \akuriyo recorded by Gildea potentially has strong \trio and/or \wayana influence \parencite[253]{gildea1998}.}
%The two forms can be reconciled by a specific idiosyncratic combination of sound changes:
%We suggest that \obj{ə} is the outcome of the lowering of the \rc{ɨ} in the prefix \rc{wɨ-}; the same has happened to the vowel in the \gl{1+2} prefix \rc{kɨt-} \exref{exp.aku-159}.
%Also, \obj{w-} appears to have been subject to ongoing erosion in \akuriyo, also evidenced by its absence in \qu{to come} and \qu{to be}, but its presence in \qu{to say} \exref{ptirw}.
%This erosion is also found in the etymologically related \gl{1}>\gl{3} prefix \exref{exp.aku-164}, as well as in other Cariban languages, like \ikpeng \pcref{sec:pekodian} or \hixka \pcref{sec:waiwaian}.
%
%
%
For \qu{to come}, \textcite[114--115]{meira1998proto} reconstructs \PTar \rc{əepɨ} for first person, and \rc{eepɨ} for the other person values, based on an idiosyncratic paradigmatic pattern in \trio and the vowel length in \akuriyo.
\akuriyo (and \carijo) then levelled this original distribution, similar to what we have suggested for Pekodian \pcref{sec:pekodian}.
We agree with this scenario, with the exception that \trio \obj{əepɨ} looks like a reflex of \rc{ət-epɨ} \pcref{sec:come}, meaning that the \PTir form would have been \rc{əʔepɨ}.

@@

In addition, \textcite{gildea1994akuriyo} recorded four more \akuriyo verbs seemingly not affected by innovative \rc{t-} \exref{akumov.aku}, all \obj{e}-initial movement verbs.
We have only found a \trio cognate for \obj{erama} \qu{to return}, which behaves like a regular \gl{s_a_} verb in taking \obj{s-} \exref{akumov.tri}.
Further, these verbs are not mentioned by \textcite{meira1998proto}, who was also working with \posscite{gildea1994akuriyo} data.
Given that this data potentially has strong \trio and/or \wayana influence \parencite[253]{gildea1998} and the lack of support by the available part of \posscite{meira1998proto} data, we cannot reconstruct these verbs as not being affected by the extension of \PTir \rc{t-}.
